Consider the following sum:
$${1\over2}+{1\over4}+{1\over8}+{1\over16}+\cdots+{1\over2^i}+\cdots$$
The dots at the end indicate that the sum goes on forever. Does this
make sense? Can we assign a numerical value to an infinite sum? While
at first it may seem difficult or impossible, we have certainly done
something similar when we talked about one quantity getting ``closer
and closer'' to a fixed quantity. Here we could ask whether, as we add
more and more terms, the sum gets closer and closer to some fixed
value. That is, look at
\begin{align*}
{1\over2}&={1\over2} \\
{3\over4}&={1\over2}+{1\over4} \\
{7\over8}&={1\over2}+{1\over4}+{1\over8} \\
{15\over16}&={1\over2}+{1\over4}+{1\over8}+{1\over16} \\
\end{align*}
and so on, and ask whether these values have a limit. They do; the limit is $1$. In fact, as we will see,
$${1\over2}+{1\over4}+{1\over8}+{1\over16}+\cdots+{1\over2^i}=
{2^i-1\over2^i}=1-{1\over2^i}$$
and then
$$\lim_{i\to\infty} 1-{1\over2^i}=1-0=1.$$

This is less ridiculous than it appears at first.  In fact, you might already believe that
$$0.3333\bar3 =
{3\over10}+{3\over100}+{3\over1000}+{3\over10000}+\cdots=
{1\over3},$$
which is similar to the sum above, except with powers of ten instead of powers of two.
And this sort of thinking is needed to make sense of numbers like $\pi$, considering
$$3.14159\ldots = 3+{1\over10}+{4\over100}+{1\over1000}+{5\over10000}+
{9\over100000}+\cdots = \pi.$$

Before we investigate infinite sums---usually called 
{\dfont series\index{series}\/}---we will investigate limits of {\dfont
  sequences\index{sequence}\/} of numbers. That is, we officially
call
$$\sum_{i=1}^\infty {1\over2^i}=
{1\over2}+{1\over4}+{1\over8}+{1\over16}+\cdots+{1\over2^i}+\cdots$$
a series, while
$${1\over2},\quad {3\over4},\quad {7\over8},\quad {15\over16},\quad \ldots,\quad {2^i-1\over2^i},\quad \ldots$$
is a sequence.  The value of a series is the limit of a particular sequence, that is,
$$\sum_{i=1}^\infty {1\over2^i}=\lim_{i\to\infty} {2^i-1\over2^i}.$$
If this all seems too obvious, let me assure you that there are twists and turns aplenty.  And if this all seems too complicated, let me assure you that we'll be going over this in much greater detail in the coming weeks. In either case, I hope that you'll join us on our journey.
